The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-05-23 Presentation/slides.qmd
updates to slides

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
subtitle: "Dissertation Defense"
format: 
  html: default
    #toc: true
    #shift-heading-level-by: 1
    #toc-expand: true
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    output-file: slides_revealjs
    slideNumber: true
    center: true
    chalkboard:
      theme: whiteboard
  #   scrollable: true
  # pptx:
  #   incremental: true
  # hugo-md:
  #   include: false
  #   html-math-method: mathjax
  #   output-file: slides_hugo.md
  gfm:
    output: false
    output-file: slides_gfm.md
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
---

::: {.content-visible when-format="html" unless-format="revealjs"}
\
[Click here](../paper.qmd) to open the dissertation manuscript.\
\
\
\
\
\
\
\
\
\

@@

<p class="buttons" style="text-align:center;">

<a class="btn btn-danger" target="_blank" href="/Presentation/slides_revealjs.html"><i class="fa-solid fa-arrow-up-right-from-square"></i> Open slides in new window </a>

</p>

::: column-body-outset-right
::: {.ratio .ratio-16x9}
<iframe src="/Presentation/slides_revealjs.html">

</iframe>
:::
:::

\
\
\
\
:::

@@

@@

![](assets/fig-toy-model1-1-1.png)

@@

## Project 1 - Methodology

-   Task description and experimental design
-   Participants and procedures
-   IGAS model introduction

![](assets/methodsFig1.png)

@@

## Project 1 - Results and Discussion

@@

## Computational Model

**Equation 1:** $$ Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}}) $$

**Equation 2:** $$ d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 } $$

@@

## Project 2 - Hit The Wall (HTW)

@@

@@

-   Task description and experimental design
-   Participants and procedures
-   Application of ALM and EXAM models

@@

## Project 2 - Results and Discussion

-   Training variability's mixed results
-   Modeling analysis with ALM and EXAM
-   Theoretical implications

@@

@@

| Dimension                                | HTT (Project 1)                                                                                                                               | HTW (Project 2)                                                                                                |
|------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------|
| Task Description                         | Projectile launching to hit a target                                                                                                          | Projectile launching to hit wall at a specific velocity                                                        |
| Task Complexity                          | More complex parabolic trajectory, both x and y velocities relevant to outcome                                                                | Simpler 1D mapping of force to outcome. Only x velocity is relevant.                                           |
| Task Space                               | More complex: xy velocity combinations closer to the solution space may still result in worse feedback due to striking the barrier.           | Simpler: smooth, linear mapping between velocity and feedback.                                                 |
| Perceptual salience of Varied Conditions | Varied conditions (# of throwing distances) are perceptually distinct, i.e. salient differences in distance between launching box and target. | Varied conditions (# of velocity bands) are less salient - only difference is the numeral displayed on screen. |
| Testing Feedback                         | Testing always included feedback                                                                                                              | Primary testing stage had no feedback.                                                                         |
| Potential for Learning during Testing    | Limited potential for learning during testing due to feedback.                                                                                | Some potential for learning during no-feedback testing by observing ball trajectory.                           |
| Training Experience                      | Varied group gets half as much experience on any one position as the constant group.                                                          | Varied group gets 1/3 as much experience on any one velocity band as the constant group.                       |
| Testing Structure                        | Random interleaving of trained/transfer testing distances.                                                                                    | Blocked structure, separately testing trained vs extrapolation testing bands.                                  |

@@

::: columns
::: {.column width="50%"}
### Theoretical Extensions

-   Investigate the role of individual differences in training variability effects.
-   Explore the impact of different types of feedback on learning and generalization.
-   Examine the influence of task complexity on the benefits of varied training.

### Methodological Innovations

-   Develop new experimental paradigms to test training variability in more complex tasks.
-   Utilize advanced computational models to simulate learning processes under varied conditions.
-   Implement longitudinal studies to assess long-term effects of training variability.
:::

::: {.column width="50%"}
### Practical Applications

-   Design adaptive training programs that adjust variability based on learner performance.
-   Apply findings to educational settings to enhance learning outcomes.
-   Use insights to improve training protocols in professional skill development.

### Collaborative Opportunities

-   Partner with educational institutions to test training variability in classroom settings.
-   Collaborate with industry professionals to apply research findings in real-world training programs.
-   Engage with interdisciplinary teams to explore the broader implications of training variability.
:::
:::

>>>>

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
subtitle: "Dissertation Defense"
format: 
  html: default
    #toc: true
    #shift-heading-level-by: 1
    #toc-expand: true
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    output-file: slides_revealjs
    slideNumber: true
    hash-type: number # Use slide numbers in the URL hash
    center: true
    scrollable: true
    chalkboard:
      theme: whiteboard
  pptx:
    incremental: true
  # hugo-md:
  #   include: false
  #   html-math-method: mathjax
  #   output-file: slides_hugo.md
  gfm:
    output: false
    output-file: slides_gfm.md
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
params:
  start_slide: 0
#engine: knitr
---

::::: {.content-visible when-format="html" unless-format="revealjs"}
\
[Click here](../paper.qmd) to open the dissertation manuscript.\
\
\
\

\

@@

```{=html}
<p class="buttons" style="text-align:center;">
<a class="btn btn-danger" target="_blank" href="/Presentation/slides_revealjs.html#{{< meta params.start_slide >}}"><i class="fa-solid fa-arrow-up-right-from-square"></i> Open slides in new window </a>
</p>
```

:::: column-body-outset-right
::: {.ratio .ratio-4x3}
```{=html}
<iframe src="/Presentation/slides_revealjs.html#{{< meta params.start_slide >}}">
</iframe>
```
:::
::::

\
\
\
:::::

@@

## Common Empirical Patterns

:::::::::  columns
::: {.column width="60%"}
### Training
- Both training conditions complete the same number of training trials.
- Varied group has worse training performance.

![](assets/train_pattern.png)
:::

::: {.column width="40%"} 
::: {.fragment}
### Testing

- Tested from novel conditions.
- Varied group has better test performance

![](assets/test_pattern.png)
:::
:::
:::::::::

## But also plenty of contradictory results and complications

- Cases where varied training makes no difference
- Cases where more training variation results in worse outcomes
- Cases where the influence of variation interacts with some other factor
    - difficulty
    - prior knowledge
    - Frequency effects, or amount of training/learning before testing

@@

<!-- ![](assets/fig-toy-model1-1-1.png) -->

@@

## Project 1 - Task & Procedure

:::: {.columns}
::: {.column width="50%"}
![](assets/methodsFig1.png)
:::

::: {.column width="50%"}
- **Training Stage** - 200 training trials. Constant groups throws from single position. Varied group 60 trials from 2 locations. 
- **Transfer Stage** - All subjects tested from both positions they were trained, and the positions trained by other group
- **Data recorded** - For every throw, recorded the X velocity and Y velocity of ball at release
:::

::::

@@

## Project 1 - Manipulations

:::: {.columns}
::: {.column width="50%"}

### Experiment 1
- Train a varied group from 2 positions, and a constant group from 1. Then test both groups, from the training positions and a novel position.

:::

::: {.column width="50%"}
### Experiment 2
- 6 constant conditions, each trained from a unique location, and a varied condition trained from 2 locations
:::

::::

## Project 1 - Results

:::: {.columns}
::: {.column width="40%"}
### Experiment 1
![](assets/fig-IGAS_Testing1-1-2.png)
:::
::: {.column width="60%"}
### Experiment 2
![](assets/fig-e2testa-1-2.png)
:::
::::

## Project 1 - Discussion

@@

## Project 1 Computational Model {.smaller .r-fit-text}

### Model of Similarity between training and solutions

![](assets/fig-taskSpace-1-1.png)

@@

## Project 1 Computational Model {.smaller .r-fit-text}

 - For each subject, compute similarity between training throws and each of the 6 testing locations
 - Test whether similarity explains the difference in performance between Constant and Varied conditions
   - Assuming equivalent generalization (1 c for both groups) - similarity improves model fit, but does not explain group difference
   - Assume training condition influences generalization gradient (fit c separately) - now similarity does explain the difference between Constant and Varied.

::: footer
**Equation 1:** $$ Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}}) $$
**Equation 2:** $$ d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 } $$
:::

## Project 1 General Discussion {.smaller .r-fit-text}

## Project 2 - Variability and Extrapolation in a Function Learning Task

@@

## Project 2 - Questions and Goals

**Empirical**
- Design a task-space large enough to assess multiple degrees of extrapolation 
- Compare varied and constant generalization from several distinct distances from their nearest training condition

**Model-based**
- If variation does influence extrapolation, can a connectionist model with similarity-based activation provide a good account?
- Can our modelling framework simultaneously account for both training and testing data?
- Accounting for the full distribution of responses

@@

-   A total of 156 participants from IU's online subject pool
-   Varied group trains from 3 "velocity bands", constant group from 1

![](assets/htw_task_fig.png)

@@

## Project 2 - Experiment 1 Results

![](assets/fig-e1-train-dev-1-1.png)

## Project 2 - Experiment 1 Results

![](assets/fig-e1-test-dev-1-3.png)

## Project 2 - Experiment 2 Design

![](assets/e2_design.png)

## Project 2 - Experiment 2 Results

![](assets/fig-e2-train-dev-1-1.png)

## Project 2 - Experiment 2 Results

![](assets/fig-e2-test-dev-1-3.png)

## Project 2 - Experiment 2 Results

![](assets/fig-e2-bmm-vx-1-2.png)

## Project 2 - Experiment 3 Design

- Ordinal feedback
- Include both the original order (E1) and reverse order (E2)

## Project 3 - Experiment 3 Results

![](assets/fig-e3-train-dev-1-1.png)

## Project 3 - Experiment 3 Results

![](assets/fig-e3-test-dev-1-3.png)

## Project 3 - Experiment 3 Results

![](assets/fig-e3-bmm-vx-1-2.png)

@@

## ALM

|                    | **ALM Response Generation**                                        |                                                                                               |
|------------------|-----------------------------|-------------------------|
| Input Activation   | $a_i(X) = \frac{e^{-c(X-X_i)^2}}{\sum_{k=1}^M e^{-c(X-X_k)^2}}$    | Input nodes activate as a function of Gaussian similarity to stimulus                         |
| Output Activation  | $O_j(X) = \sum_{k=1}^M w_{ji} \cdot a_i(X)$                        | Output unit $O_j$ activation is the weighted sum of input activations and association weights |
| Output Probability | $P[Y_j|X] = \frac{O_j(X)}{\sum_{k=1}^M O_k(X)}$                    | The response, $Y_j$ probabilites computed via Luce's choice rule                              |
| Mean Output        | $m(X) = \sum_{j=1}^L Y_j \cdot \frac{O_j(x)}{\sum_{k=1}^M O_k(X)}$ | Weighted average of probabilities determines response to X                                    |
|                    | **ALM Learning**                                                   |                                                                                               |
| Feedback           | $f_j(Z) = e^{-c(Z-Y_j)^2}$                                         | feedback signal Z computed as similarity between ideal response and observed response         |
| magnitude of error | $\Delta_{ji}=(f_{j}(Z)-o_{j}(X))a_{i}(X)$                          | Delta rule to update weights.                                                                 |
| Update Weights     | $w_{ji}^{new}=w_{ji}+\eta\Delta_{ji}$                              | Updates scaled by learning rate parameter $\eta$.                                             |

## EXAM

|                    | **EXAM Response Generation**                                                                                     |                                                                                                               |
|--------------------|-----------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------|
| Instance Retrieval | $P[X_i|X] = \frac{a_i(X)}{\sum_{k=1}^M a_k(X)}$                                                                 | Novel test stimulus $X$ activates input nodes $X_i$                                                           |
| Slope Computation  | $S = \frac{m(X_{1})-m(X_{2})}{X_{1}-X_{2}}$                                                                     | Slope value, $S$ computed from nearest training instances                                                     |
| Response           | $E[Y|X_i] = m(X_i) + S \cdot [X - X_i]$                                                                         | Final EXAM response is the ALM response for the nearest training stimulus, $m(X_i)$, adjusted by local slope $S$. |

## Project 2 - Model Fitting Procedure

### Approximate Bayesian Computation

## Modelling Results

![](assets/fig-ee-e1-1.png)

@@

| Dimension | HTT (Project 1) | HTW (Project 2) |
|----------------|--------------------------------|-------------------------|
| Task Description | Projectile launching to hit a target | Projectile launching to hit wall at a specific velocity |
| Task Complexity | More complex parabolic trajectory, both x and y velocities relevant to outcome | Simpler 1D mapping of force to outcome. Only x velocity is relevant. |
| Task Space | More complex: xy velocity combinations closer to the solution space may still result in worse feedback due to striking the barrier. | Simpler: smooth, linear mapping between velocity and feedback. |
| Perceptual salience of Varied Conditions | Varied conditions (# of throwing distances) are perceptually distinct, i.e. salient differences in distance between launching box and target. | Varied conditions (# of velocity bands) are less salient - only difference is the numeral displayed on screen. |
| Testing Feedback | Testing always included feedback | Primary testing stage had no feedback. |
| Potential for Learning during Testing | Limited potential for learning during testing due to feedback. | Some potential for learning during no-feedback testing by observing ball trajectory. |
| Training Experience | Varied group gets half as much experience on any one position as the constant group. | Varied group gets 1/3 as much experience on any one velocity band as the constant group. |
| Testing Structure | Random interleaving of trained/transfer testing distances. | Blocked structure, separately testing trained vs extrapolation testing bands. |

@@

::::::::: columns
::::: {.column width="50%"}
::: fragment
### Theoretical Extensions

-   Investigate the role of individual differences in training variability effects.
-   Explore the impact of different types of feedback on learning and generalization.
-   Examine the influence of task complexity on the benefits of varied training.
:::

::: fragment
### Methodological Innovations

-   Develop new experimental paradigms to test training variability in more complex tasks.
-   Utilize advanced computational models to simulate learning processes under varied conditions.
-   Implement longitudinal studies to assess long-term effects of training variability.
:::
:::::

::::: {.column width="50%"}
::: fragment
### Practical Applications

-   Design adaptive training programs that adjust variability based on learner performance.
-   Apply findings to educational settings to enhance learning outcomes.
-   Use insights to improve training protocols in professional skill development.
:::

::: fragment
### Collaborative Opportunities

-   Partner with educational institutions to test training variability in classroom settings.
-   Collaborate with industry professionals to apply research findings in real-world training programs.
-   Engage with interdisciplinary teams to explore the broader implications of training variability.
:::
:::::
:::::::::
